Modelling circumstellar discs with 3D radiation hydrodynamics

### 2009-10-09 SPH-Torus-paper1/sphtorus_benchmark.tex
Updates following meeting of DA and TH.
git-svn-id: https://repository.astro.ex.ac.uk/torus/trunk@2047 25464f21-6099-402c-b30a-640f1d40d38b

In figure xxx we plot the temperature distribution at radiative equilibrium for the benchmark disc. 
{\sc torus} calculates the temperature distribution
using a high resolution 2D geometry, in order to determine the correct
temperature as a function of cylindrical polar $r$ and $z$
co-ordinates. This is then compared to the temperature from the SPH
disc for each SPH particle. Positive values of the fractional error
indicate that the SPH disc is hotter than the expected value. The temperatures show deviations of $\pm$ 50\% (XXXX CHECK) from the benchmark values, a deviation that is purely driven by the resolution of the density field derived from the SPH particles, since a `{\sc torus} only' calculation, where the AMR grid is refined on optical depth criteria and the cell densities are found from the analytical description of the density structure (equation xxxx), matches the benchmark temperatures to within 2\%. The systematic differences in the temperature distribution are due to an increased vertical transport of radiation from the surface layers of the disc to the mid-plane due to a lack of spatial resolution, leading to errors in the calculation of the specific intensities (particularly in the inner disc region).  These systematic differences are more readily apparent in the emergent SEDs, which are a more sensitive probe of the accuracy of the RT calculation than  the temperature distribution.

@@

SEDs are plotted in Fig.~\ref{fig:n_part_sed_013} for a viewing angle
of 12.5 degrees and Fig.~\ref{fig:n_part_sed_077} for a viewing angle
of 77.5 degrees. At the shortest wavelengths the 12.5 degrees
inclination angle SED is dominated by the photosphere of the central
star. This region of the SED is well represented with $10^5$ and
$10^8$ particles but is of course less well represented with $10^6$ or $10^7$
particles where the central star is obscured. At the longest
wavelengths the disc is optically thin and the SEDs will be well
modelled provided the total disc mass is accurately represented. In
all four cases the total disc mass is well represented (XXX cite section) and the long
wavelength SED matches the benchmark results. At intermediate
wavelengths the SED is sensitive to the details of the disc
representation and it is clear that accurately modelling the SED from
this disc is challenging, even with $10^8$ particles. (XXXX need to discuss this in the conclusion).
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e5}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e6}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e7}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e8}}
  \caption{SEDs at a 12.5 degree inclination angle for discs represented by different numbers of SPH
    particles. The benchmark result is plotted as a solid line and the
  modelled SED is plotted as a dashed line. } 
  \label{fig:n_part_sed_013}
\end{figure*}
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e5}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e6}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e7}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e8}}
  \caption{SEDs at a 77.5 degree inclination angle for discs represented by different numbers of SPH
    particles. The benchmark result is plotted as a solid line and the
  modelled SED is plotted as a dashed line. } 
  \label{fig:n_part_sed_077}
\end{figure*}

@@

The frequency with which the radiative transfer is calculated is determined by the
sound crossing time in the inner regions of the disc as this is the
time scale over which the scale height in this region will vary and
effect the transfer of radiation into the outer parts of the disc. (XXX put numbers
here to justify the 1 every 4). A radiative transfer calculation is performed after every four
hydrodynamics time steps (equivalent to every 2.4 years).

@@

Transient oscillations, as the disc adjusts towards equilibrium, are
observed whether radiative transfer is included or not. They are
vertical motions which decay more rapidly at smaller radii and hence
give the appearance of propagating outwards, even though the motion is
vertical. Other disc models using radiation hydrodynamics have found
oscillatory solutions with inwards propagating waves
(e.g. \cite{min_2009}) which are fundamentally different to the
transient effects seen in this case. However it should be noted that
\cite{min_2009} only found an oscillatory solution in their most
optically thick disc, which is significantly more massive than our
model disc, and may represent a different regime to that considered
here. We note that the disc retains its azimuthal symmetry throughout 
(XXXX Check!) the simulation and we find no evidence for warping instabilities.

@@

Although we have found a method for constructing an AMR grid from SPH
particles which is robust, in terms of the total mass on the grid, we
found that it was necessary to add additional grid refinement in the
central region of the disc to allow the inner edge to be well
represented. The inner edge of the disc has a large
effect on the temperature distribution in the disc and on the emergent
SEDs. Consequently the accuracy of the solution is dependent upon the
accurate representation of a region which constitutes a small fraction
of the disc by volume. As the method combines both SPH and grid
elements it is important that both these components have sufficient
resolution to represent the disc inner edge. This presents a
particular challenge for a SPH code which uses equal mass particles
but is achievable with a large number of particles. Accurately
modelling the SED using this method seems to be considerably more
demanding than accurately representing the temperature structure
within the disc and temperature errors can be reduced to a manageable
level with a realistic number of SPH particles in the disc.

>>>>

In Fig.~\ref{fig:nomod_tem_diff} we plot the fractional error in the
temperature distribution at radiative equilibrium for the benchmark
disc.  {\sc torus} calculates the temperature distribution using a
high resolution 2D geometry, in order to determine the correct
temperature as a function of cylindrical polar $r$ and $z$
co-ordinates. This is then compared to the temperature from the SPH
disc for each SPH particle. Positive values of the fractional error
indicate that the SPH disc is hotter than the expected value. The
temperatures show deviations of $-30$\% to $+50$\%  from the
benchmark values, a deviation that is purely driven by the resolution
of the density field derived from the SPH particles, since a `{\sc
  torus} only' calculation, where the AMR grid is refined on optical
depth criteria and the cell densities are found from the analytical
description of the density structure (Eqn.~\ref{eqn:pascucci_density}), matches the
benchmark temperatures to within 2\%. The systematic differences in
the temperature distribution are due to an increased vertical
transport of radiation from the surface layers of the disc to the
mid-plane due to a lack of spatial resolution, leading to errors in
the calculation of the specific intensities (particularly in the inner
disc region).  These systematic differences are more readily apparent
in the emergent SEDs, which are a more sensitive probe of the accuracy
of the RT calculation than the temperature distribution.

\begin{figure}
  \includegraphics[scale=0.3]{figures/nomod_tem_diff}
  \caption{Temperature difference between SPH benchmark disc and high
    accuracy 2D result.}
  \label{fig:nomod_tem_diff}
\end{figure}

@@

SEDs are plotted in Fig.~\ref{fig:n_part_sed_013} for a viewing angle
of 12.5 degrees and Fig.~\ref{fig:n_part_sed_077} for a viewing angle
of 77.5 degrees. At the shortest wavelengths the 12.5 degrees
inclination angle SED is dominated by the photosphere of the central
star. This region of the SED is well represented with $10^5$ and
$10^8$ particles but is of course less well represented with $10^6$ or $10^7$
particles where the central star is obscured. At the longest
wavelengths the disc is optically thin and the SEDs will be well
modelled provided the total disc mass is accurately represented. In
all four cases the total disc mass is well represented (see Section~\ref{section:grid_generation}) and the long
wavelength SED matches the benchmark results. At intermediate
wavelengths the SED is sensitive to the details of the disc
representation and it is clear that accurately modelling the SED from
this disc is challenging, even with $10^8$ particles.
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e5}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e6}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e7}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e8}}
  \caption{SEDs at a 12.5 degree inclination angle for discs represented by different numbers of SPH
    particles. The benchmark result is plotted as a solid line and the
  modelled SED is plotted as a dashed line. } 
  \label{fig:n_part_sed_013}
\end{figure*}
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e5}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e6}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e7}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e8}}
  \caption{SEDs at a 77.5 degree inclination angle for discs represented by different numbers of SPH
    particles. The benchmark result is plotted as a solid line and the
  modelled SED is plotted as a dashed line. } 
  \label{fig:n_part_sed_077}
\end{figure*}

@@

The frequency with which the radiative transfer is calculated is
determined by the sound crossing time in the inner regions of the disc
as this is the time scale over which the scale height in this region
will vary and effect the transfer of radiation into the outer parts of
the disc.  The vertical sound crossing time at a radius of
10~au is typically 2 years, so a radiative transfer calculation is
performed after every four hydrodynamics time steps (equivalent to
every 2.4 years).

@@

Transient oscillations, as the disc adjusts towards equilibrium, are
observed whether radiative transfer is included or not. They are
vertical motions which decay more rapidly at smaller radii and hence
give the appearance of propagating outwards, even though the motion is
vertical. Other disc models using radiation hydrodynamics have found
oscillatory solutions with inwards propagating waves
(e.g. \cite{min_2009}) which are fundamentally different to the
transient effects seen in this case. However it should be noted that
\cite{min_2009} only found an oscillatory solution in their most
optically thick disc, which is significantly more massive than our
model disc, and may represent a different regime to that considered
here. We note that the disc retains its azimuthal symmetry throughout 
the simulation and we find no evidence for warping instabilities.

@@

Although we have found a method for constructing an AMR grid from SPH
particles which is robust, in terms of the total mass on the grid, we
found that it was necessary to add additional grid refinement in the
central region of the disc to allow the inner edge to be well
represented. The inner edge of the disc has a large
effect on the temperature distribution in the disc and on the emergent
SEDs. Consequently the accuracy of the solution is dependent upon the
accurate representation of a region which constitutes a small fraction
of the disc by volume. As the method combines both SPH and grid
elements it is important that both these components have sufficient
resolution to represent the disc inner edge. This presents a
particular challenge for a SPH code which uses equal mass particles
but is achievable with a large number of particles. Accurately
modelling the SED using this method is considerably more
demanding than accurately representing the temperature structure
within the disc. Temperature errors can be reduced to a manageable
level using a realistic number of SPH particles, yielding a 
temperature distribution suitable for use in a hydrodynamics
calculation, however even with a disc composed of $10^8$ SPH particles
the modelled SEDs show significant deviations from the benchmark results.
